MS_Intro_Short

### 2024-02-16 Markdown/Manuscript/Child_RMDs/Clippings.Rmd
minor edits to body

Computational models, particularly those addressing economic topics, have become essential tools for health policy development [@dakin2015influence; @Erdemir2020]. Although influential and widely used, health economic models typically have a number of limitations that at best restrict their usefulness and which in some cases may facilitate harmful misuse.

@@

The resource and skills intensive undertaking of modelling could be made more tractable if it was easier for health economists to re-use each other's models [@Arnold2010]. However, as many health economic models are owned by pharmaceutical companies and consultancies, commercial considerations can limit the reuse of models and their constituent code and data [@Feenstra2022]. Legal and ethical issues such as privacy and confidentiality requirements further limit public release of some model artefacts [@Pouwels2022]. Transferring a health economic model developed for one jurisdiction for application in another typically involves retaining some features and updating others [@barbieri2010international]. However, models do not always make implementation choices (for example, use of concepts with standardised meanings across jurisdictions<!-- [@garcia2021cost] -->, supplying context specific data as replaceable data-packs and distributing source code under licenses that allow derivative works) that facilitate model transferability.

@@

A potential strategy for improving model accountability, reusability and updatability is to make models both modular and open-source.

A modular model is constructed from multiple reusable and replaceable sub-models (modules) [@pan2021modular]. Modules can share inputs and outputs with other modules or can be run as independent models [@barros2023empowering]. Advantages of modular models include feasibility (large projects are broken into smaller tasks, with each component independently developed and tested) and flexibility (making it easier to selectively replace or update specific parts of a model and to scale up or down the level of granularity) [@pan2021modular]. Modular approaches are currently being used to facilitate the development of complex computational models in disciplines such as biology [@pan2021modular], ecology [@barros2023empowering] and neuroscience [@frazier2022robust]. In health economics the related and enabling concept of reference models has been recommended [@Afzali2013], but peer reviewed studies describing modular health economic models remain relatively rare, though examples exist in infectious disease [@Trauer2017] and cardiology [@Urach2013].

Modular models provide an opportunity for multiple modelling teams to contribute to, test and reuse models. To enhance this capacity, modular models may be implemented as open source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust]. Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019]. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing OSHEMs remain scarce, piecemeal and need improving [@Sampson2019].

>>>>

@@

, supplying context specific data as replaceable data-packs and distributing source code under licenses that allow derivative works) that facilitate model transferability.

@@

### 2024-02-16 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
minor edits to body

(i)	proposed criteria for assessing ethical CHEM implementations;

(ii) a prototype software framework for the ethical implementation of CHEMs; and

(iii)	how the software framework is being used to develop an open-source CHEM in youth mental health, with an initial focus on valuing health outcomes.

>>>>

(i)	a prototype software framework we have developed to support the development of modular and open source CHEMs; and
<!-- propose a set of ethical responsibilities for CHEM project teams and criteria for assessing ethical CHEM implementations; -->
<!-- (ii)  -->
(ii)	how we have applied that software framework to develop and share an initial set of modules for a economic model of youth mental health.

### 2024-02-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Toolkit.Rmd
minor edits to body

To support the development of CHEMs that are Transparent, Reusable and Updatable (TRU),<!-- assessment criteria,  --> we have created a prototype software framework called ready4. The software framework aims to provide a toolkit for: (i) enabling modular implementation of CHEMs; (ii) authoring and documenting CHEM modules; (iii) managing the labelling and transfer of CHEM input and output data; and (iv) authoring reproducible analyses that apply CHEMs to compatible data.

>>>>

## Purpose
The overarching goal of our software framework, which we call ready4, is to support the development of CHEMs that meet explicit criteria for transparency, reusability and updatability. We have describe these criteria elsewhere [@ethicsCHEMS2024], but briefly they are: software files are open access, developer contributions are easily identified, programming practices promote selective reuse, licenses restrict only unethical reuse, maintenance infrastructure is in place and systematic retesting and deprecation with each new release.

## User requirements

1) Provide a standardised template model module and programming syntax.

2) Develop consolidated project documentation and provide tools to find model modules, datasets and study replication programs.

3) Author, document and share model modules, integrating with online services for open source software development

4) Ingest, label and share model data, integrating with online services for open science data archiving.

5) Implement integrated and reproducible data ingest, analysis and reporting pipelines.

## XXXYYY
The software framework aims to provide a toolkit for: (i) enabling modular implementation of CHEMs; (ii) authoring and documenting CHEM modules; (iii) managing the labelling and transfer of CHEM input and output data; and (iv) authoring reproducible analyses that apply CHEMs to compatible data.

### 2024-02-16 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
minor edits to body

@@

<!-- Better software frameworks could potentially help address some of the barriers to open-source CHEM development.  -->
A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python.

A high-level software framework for implementing open-source <!-- Modification_start -->Computational Health Economic Models (CHEMs)<!-- Modification_end --> in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, <!-- the ethical responsibilities of public health computational modellers remain poorly understood [@10.3389/fpubh.2017.00068] and  -->the computational modelling landscape is changing fast [@calder2018computational]. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software]. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development].

>>>>

<!-- Import_start -->Computational models, particularly those addressing economic topics, have become essential tools for health policy development [@dakin2015influence; @Erdemir2020]. Although influential and widely used, health economic models can be skills and resource intensive to undertake.<!-- Import_end --> One suggestion for making health economic modelling projects more tractable is for health economists to re-use each other's models [@Arnold2010]. A potential strategy for achieving this goal is to make computational health economic models (CHEMs) both modular and open-source. A modular model is constructed from multiple reusable and replaceable sub-models (modules) [@pan2021modular]. Modules can share inputs and outputs with other modules or can be run as independent models [@barros2023empowering]. Advantages of modular models include feasibility (large projects are broken into smaller tasks, with each component independently developed and tested) and flexibility (making it easier to selectively replace or update specific parts of a model and to scale up or down the level of granularity) [@pan2021modular]. Modular approaches are currently being used to facilitate the development of complex computational models in disciplines such as biology [@pan2021modular], ecology [@barros2023empowering] and neuroscience [@frazier2022robust]. In health economics the related and enabling concept of reference models has been recommended [@Afzali2013], but peer reviewed studies describing modular health economic models remain relatively rare, though examples exist in infectious disease [@Trauer2017] and cardiology [@Urach2013].

Modular CHEMs provide an opportunity for multiple modelling teams to contribute to, test and reuse models. To enhance this capacity, modular CHEMs may be implemented as open source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust]. Although there appears to be in principle support from many health economists for greater use of open source CHEMs  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of open source CHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019]. <!-- Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but  --><!-- Guidelines for implementing open source CHEMs remain scarce, piecemeal and need improving [@Sampson2019].  -->As many health economic models are owned by pharmaceutical companies and consultancies, commercial considerations can limit the reuse of models and their constituent code and data [@Feenstra2022]. Legal and ethical issues such as privacy and confidentiality requirements further limit public release of some model artefacts [@Pouwels2022].

<!-- Transferring a health economic model developed for one jurisdiction for application in another typically involves retaining some features and updating others [@barbieri2010international]. However, models do not always make implementation choices (for example, use of concepts with standardised meanings across jurisdictions)<!-- [@garcia2021cost] -->  <!-- facilitate such selective re-use. -->

@@

<!-- typically have a number of limitations that at best restrict their usefulness and which in some cases may facilitate harmful misuse. -->

Some of the barriers<!-- Give examples --> to the development of open-source and modular CHEMs may be amenable to technical solutions such as software frameworks.<!-- Check last sentence was not used in ethics manuscript  --> A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python. However, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software]. Software frameworks can also be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development].

<!-- There are multiple challenges to overcome when developing a software framework. Firstly, <!-- the ethical responsibilities of public health computational modellers remain poorly understood [@10.3389/fpubh.2017.00068] and  --><!-- the computational modelling landscape is changing fast [@calder2018computational].  -->

<!-- A high-level software framework for implementing open-source <!-- Modification_start --><!--  CHEMsin R has previously been developed with the primary aim of improving model transparency [@alarid2019need].  -->

<!-- However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements.  -->
